Final Project Report - An exploration of Biliary Cirrhosis and Treatments

### 2023-12-16 Submission.Rmd
CHANGES

---
title: 'Final Project Report - An exploration of Biliary Cirrhosis and Treatments'
output:
  pdf_document: default
authors: Mihir Kulkarni+Nithika Menon
---

@@

![Primary biliary cholangitis (https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874)](diagram.jpeg)

@@

```{r}
#?????
cirrhosis %>% 
  ggplot(aes(x = N_Days)) + geom_histogram() +
  facet_wrap(facets = vars(Status))

@@

Stepwise results:

```{r models}
# Lower bound (Intercept only)
lower <- glm(
  formula = Status ~ 1,
  data = trainingData,
  family = binomial
)

@@

# Results

@@

## Model 1

@@

@@

# Testing our model

@@

#Comparison between models

Now let us compare the performance of our logistic regression model with a decision tree model. We will use the same train test split as before.

@@

Now we can go back to our initial goal: comparing how the custers compare to the Stage variable

```{r}
cirrhosisCluster2$cluster <- hybridCirrhosis$cluster
# Calculate mean (or median, etc.) for each variable in each cluster
library(dplyr)
cluster_summary <- cirrhosisCluster2 %>%
  group_by(cluster) %>%
  summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Stages within Clusters",
       x = "Cluster",
       y = "Count",
       fill = "Stage") +
  theme_minimal()

@@

We can see that the clusters are not evenly distributed. This is a good sign, as it means that the clusters are not just a random grouping of the data.

@@

@@

# References--Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.

>>>>

---
title: 'Final Project Report - An exploration of Biliary Cirrhosis and Treatments'
output:
  pdf_document: default
  toc: true
authors: Mihir Kulkarni+Nithika Menon
---

@@

![Primary biliary cholangitis (<https://www.mayoclinic.org/diseases-conditions/primary-biliary-cholangitis/symptoms-causes/syc-20376874>)](diagram.jpeg)

@@

@@

```{r results = 'hide'}
# Lower bound (Intercept only)
lower <- glm(
  formula = Status ~ 1,
  data = trainingData,
  family = binomial
)

@@

@@

### Model 1

@@

### Model 2

@@

## Testing our model

@@

# Discussion: Comparison between models

Let us recall the differences between the two models:

Logistic Regression:

-   Pros: Simple, efficient, good for linear relationships, provides outcome probabilities

-   Cons: Assumes linear decision boundary, struggles with complex relationships, sensitive to imbalanced data

Decision Trees:

-   Pros: Handles non-linear data, easy to interpret, works with mixed data types, non-parametric, N/A values

-   Cons: Prone to overfitting, sensitive to data changes, poor at extrapolation, biased with imbalanced datasets

Now let us compare the performance of our logistic regression model with a decision tree model.

| Testing Data Score | CART-based approach | Regression based Approach |
|--------------------|---------------------|---------------------------|
| Accuracy           | 72.9%               | 71.43%                    |
| Precision          | 57.6%               | 43.48%                    |
| Recall             | 82.70%              | 76.92%                    |

We can see that for this data, surprisingly, the tree based approach had better results. This is an uncommon observation. We believe this was because of the sparcity of the data and how the rows N/A values prevented over fitting with the small data size.

@@

Now we can go back to our initial goal: comparing how the clusters compare to the Stage variable.

```{r}
cirrhosisCluster2$cluster <- hybridCirrhosis2$cluster
# Calculate mean (or median, etc.) for each variable in each cluster
library(dplyr)
cluster_summary <- cirrhosisCluster2 %>%
  group_by(cluster) %>%
  summarise_all(funs(mean(., na.rm = TRUE))) # Replace mean with median or any other function as necessary
kable(cluster_summary %>% dplyr::select(-Status, -Drug, - Sex, -Edema, -Stage), label = "Cluster Summary")
ggplot(cirrhosisCluster2, aes(x = factor(cluster), fill = factor(Stage))) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Stages within Clusters",
       x = "Cluster",
       y = "Count",
       fill = "Stage") +
  theme_minimal()

@@

We can see that the clusters have a somewhat connection to the Stage, but is not as direct as we hypothesized. We were also able to find the cluster summary compared to the rest of the variables and were able to plot that into a table.

@@

To conclude, in this report, we were able to explore Biliary Cirrhosis and Treatments through various angles. Though our clustering we were able to find new patterns among the data. Though the logistic and CART algorithms we were able to build models that were able to predict the chance of a patient's death using various predictors.

Along with the analysis in the report project, we attempted to explore various models such as predicting the likelihood of a liver transplant and the number of days until an event. However, due to our data's and report's limitations, such as the limited data quantities, we were unable to get concrete solutions. We hope that our results can be used to build even better models, making healthcare more inclusive, efficient and affordable.

@@

We collaborated throughout the document using GitHub, but here are the main responsibilities:

**Introduction: Nithika**

**EDA: Nithika**

**Tree: Mihir**

**Logistic Regression: Mihir**

**Clustering: Nithika**

**Discussion: Mihir**

# References

Dataset: <https://www.kaggle.com/datasets/fedesoriano/cirrhosis-prediction-dataset/data>

Study: <https://faculty.washington.edu/abansal/ShortCourse_DynamicDecisionMaking/Dickson1989_MayoPBCOriginalArticle.pdf>
